A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-12 inst/manuscript/manuscript.tex
section 2 updates

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A comparison of design-based and model-based approaches for spatial data.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A comparison of design-based and model-based approaches for spatial
data.}
    \author[USEPA]{In alphabetical order Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. The rest of the manuscript is organized as follows. In
Section \ref{sec:background}, we compare sampling and estimation
procedures between the design-based approach and the model-based
approach for spatial data. In Section \ref{sec:numstudy}, we use a
simulation approach to study the behavior and performance of both
approaches. In Section \ref{application}, we use both approaches to
analyze real data. And in Section \ref{sec:discussion}, we end with a
discussion and provide directions for future research.

@@

\hypertarget{comparing-design-based-and-model-based-approaches}{%
\subsection{Comparing Design-Based and Model-Based
Approaches}\label{comparing-design-based-and-model-based-approaches}}

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sample, and cluster sampling. These sampling
designs tend to select units from the sampling frame independently of
other units, so we call them ``Independent Random Sampling'' (IRS)
designs. The design-based approach combines the randomness of the
sampling design and the data collected via the sample to estimate
parameters (e.g.~means and totals) of a population. Generally, these
population parameters are assumed to be fixed, unknown constants.

@@

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is often incorporated through
distributional assumptions on this process and need not be incorporated
through random sampling (bottom row of Figure \ref{fig:fig1}). Instead
of estimating fixed but unknown parameters (as in the design-based
approach), the goal of model-based inference in the spatial context is
often to predict a realized variable, or value. For example, suppose the
realized mean of all population units is the value of interest. Instead
of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units (and not any units outside the population), we would be
able to exactly predict (without any uncertainty) the mean of our one
realized process. Assuming the data are realized from a specific
data-generating process yields predictions that are linked to
distributional assumptions. These distributional assumptions are used to
derive prediction intervals. If we repeatedly generate the response
values from a fixed spatial process and obtained a sample, then 95\% of
all 95\% prediction intervals constructed from a procedure with
appropriate coverage will contain their respective realized means.
Cressie (1993) and Schabenberger and Gotway (2017) provide reviews of
model-based approaches for spatial data. A visual comparison of the
design-based and model-based assumptinos is provided in Figure
\ref{fig:fig1} (Brus (2020) provides a similar figure).

@@

\textbf{Tony O.}: Before this section is it useful to have a section
that lays out the general site selection and general analysis options.
Thinking about site selection as design-based IRS, design-based GRTS,
Arbitrary set of sites, selection for model-based. Then general analysis
options as design-based no spatial, design-based spatial, model-based.
This four by three table would show that model-based analyses are
possible for all selection options. Design-based options with no spatial
info possible for IRS-based and GRTS-based. Design-based options with
spatial info possible for GRTS-based.

\textbf{Jay VH}: What about the design for model-based inference?
Strictly speaking, it is fixed -- there is no probabilistic use of a
randomized design. However, we are going to have to deal with Diggle et
al.~(2010).

\hypertarget{spatially-balanced-design-and-analysis}{%
\subsection{Spatially Balanced Design and
Analysis}\label{spatially-balanced-design-and-analysis}}

\textbf{Lisa M.}: Need a more precise definition of ``miniature'' in
this context, and need an example.

\textbf{Jay VH}: Saying ``the distribution of the sampled population
units mirrors the density of\ldots{}'' is confusing to me. Are these
formal statistical definitions of distribution (cumulative distribution
function) and density (probability density function)? Wouldn't IRS
sample be a miniature, as it should, on average, mirror a population?

The design-based approach can use spatial locations to obtain spatially
balanced samples. First we discuss spatial balance with respect to the
population (Stevens and Olsen, 2004). A sample is spatially balanced
with respect to the population if the sampled population units are a
miniature of the population units. A sample is a miniature of the
population if the distribution of the sampled population units mirrors
the density of all population units. Spatial balance with respect to the
population is different than spatial balance with respect to geography.
A sample that is spatially balanced with respect to geography is spread
out in some type of equidistant manner over geographical space and is
not meant to be miniatures of the population. When we refer to spatial
balance henceforth, we mean spatial balance with respect to the
population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens and Olsen, 2004; Wang et al., 2013). To quantify spatial
balance, Stevens and Olsen (2004) proposed loss functions based on
Voroni polygons. The first spatially balanced sampling algorithm that
saw widespread use was the Generalized Random Tessellation Stratified
(Stevens and Olsen, 2004). Since GRTS was developed, several other
spatially balanced sampling algorithms have emerged, including the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018). We focus on the Generalized Random
Tessellation Stratified (GRTS) algorithm to select spatially balanced
sampling because it has several attractive properties, including
\textbf{Lisa M.}: List major attractive properties, and detailed by
Stevens and Olsen (2004) and Dumelle et al. (2021).

The GRTS algorithm is used to sample from finite and infinite
populations and works by utilizing a mapping between two-dimensional and
one-dimensional space. The population units in two-dimensional space are
divided into cells using a hierarchical index. Population units are then
mapped to a one-dimensional line via the hierarchical indexing. The line
length of each population unit equals its inclusion probability. A
systematic sample is conducted on the line and these samples are linked
to a population unit in two-dimensional space, which results in the
desired sample. Stevens and Olsen (2004) and Dumelle et al. (2021)
provide further details.

After collecting a sample using the GRTS algorithm, the data are used to
estimate population parameters. The Horvitz-Thompson estimator (Horvitz
and Thompson, 1952) yields unbiased estimates of population means and
totals. For example, if \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\) (denoted by \(\hat{\tau}_{ht}\)),
is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(\pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. A similar formula exists for estimating the mean, \(\mu\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but they have two drawbacks. First, they rely
on calculating \(\pi_{ij}\), the probability that population unit \(i\)
and population unit \(j\) are included in the sample, and this can be
very difficult to calculate. Second, they ignore the spatial locations
of the population units. To address these drawbacks, Stevens and Olsen
(2003) proposed a local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\), and it
incorporates spatial locations by assigning higher weights to nearby
observations. Stevens and Olsen (2003) show this variance estimators
tends to reduce the estimated standard error of \(\hat{\tau}\), yielding
narrower confidence confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of basing inference off of a specific
sampling design, we assume the data are generated by a spatial process.
Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. Let
\(\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}\)
be a response vector at locations \(s_1\), \(s_2\), . . . , \(s_N\) that
can be measured at the \(N\) population units and is represented as an
\(N \times 1\) vector. Suppose we want to predict some linear function
of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

However, we often only have a sample of the \(N\) population units.
Denoting quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with a subscript \emph{u},

@@

We also assume that there is spatial correlation in \(\bm{\delta}\),
which can be modeled using a covariance function. It is common to assume
the covariance function is second-order stationary and isotropic
(Cressie, 1993), and that the spatial covariance decreases as the
separation between population units increases. Many spatial covariance
functions exist, but the primary function we use throughout the
simulations and applications in this manuscript is the exponential
covariance function: the \(i,j^{th}\) entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align} where \(h_{i,j}\) is the distance between population units
\(i\) and \(j\), and \(\bm{\theta}\) is a vector of spatial covariance
parameters of the partial sill \(\theta_1\), the range \(\theta_2\), and
the nugget \(\theta_3\); and, \(\mathbbm{1}\) is equal to 1 when
distance \(h_i,j\) is equal to 0, and equal to 0 otherwise. However, any
spatial covariance function could be used in the place of the
exponential, including functions that allow for non-stationarity or
anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

\textbf{Lisa M.} : Include formulas. Perhaps, but, these are very heavy
in notation and matrix algebra. We might consider, however, adding the
formulas to an Appendix.

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in (Ver Hoef, 2008), we
note here that the predictor and its variance are both moment-based.

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses in a finite population
setting. We choose to use FPBK because it is faster than a Bayesian
approach and random forest and because Ver Hoef and Temesgen (2013)
showed that the method outperforms k-nearest-neighbors in many
scenarios.

@@

\begin{longtable}[]{@{}lll@{}}
\caption{\label{tab:designanalysis} Types of Sampling Design and
Analysis combinations considered in the simulation study. The columns
give the two types of sampling designs while the rows give the two types
of analyses.}\tabularnewline
\toprule
& IRS & GRTS\tabularnewline
\midrule
\endfirsthead
\toprule
& IRS & GRTS\tabularnewline
\midrule
\endhead
Design & IRS-Design & GRTS-Design\tabularnewline
Model & IRS-Model & GRTS-Model\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[]{@{}llll@{}}
\caption{\label{tab:parmtab} Simulation parameters. Total variability
for all scenarios was 2 so that the partial sill was 0, 1, or
1.8.}\tabularnewline
\toprule
\endhead
Sample Size (n) & 50 & 100 & 200\tabularnewline
Site Locations & Random & Gridded &\tabularnewline
Partial Sill / Total Variance & 0 & 0.5 & 0.9\tabularnewline
Response Type & Normal & Lognormal &\tabularnewline
\bottomrule
\end{longtable}

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A comparison of design-based and model-based approaches for finite population spatial data.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A comparison of design-based and model-based approaches for finite
population spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. Though these comparisons generalize to both finite
populations (e.g.~point resources) and infinite populations (e.g.~linear
and areal resources), we focus on applications to finite populations.
The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach for
spatial data. In Section \ref{sec:numstudy}, we use a simulation
approach to study the behavior and performance of both approaches. In
Section \ref{application}, we use both approaches to analyze real data.
And in Section \ref{sec:discussion}, we end with a discussion and
provide directions for future research.

@@

\hypertarget{subsec:dvm_compare}{%
\subsection{Comparing Design-Based and Model-Based
Approaches}\label{subsec:dvm_compare}}

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sample, and cluster sampling. These sampling
designs tend to select units from the sampling frame independently of
other units, so we call them ``Independent Random Sampling'' (IRS)
designs. Sampling designs incorporating the spatial locations of units
in the sample frame are called spatially balanced designs. Spatially
balanced designs can be obtained using the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004),
which we discuss in more detail in Section \ref{subsec:spb_design}. The
design-based approach combines the randomness of the sampling design and
the data collected via the sample to estimate parameters (e.g.~means and
totals) of a population. Generally, these population parameters are
assumed to be fixed, unknown constants.

@@

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement random sampling.

Instead of estimating fixed but unknown parameters (as in the
design-based approach), the goal of model-based inference in the spatial
context is often to predict a realized variable, or value. For example,
suppose the realized mean of all population units is the value of
interest. Instead of \emph{estimating} a fixed, unknown mean, we are
\emph{predicting} the value of the mean, a random variable. Prediction
intervals are then derived leveraging assumptions of the data generating
process. If we repeatedly generate the response values from a fixed
spatial process and obtained a sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Brus
(2020) provides a similar figure).

@@

\hypertarget{subsec:spb_design}{%
\subsection{Spatially Balanced Design and
Analysis}\label{subsec:spb_design}}

Spatially balanced samples can be obtained using the design-based
approach. Spatially balanced samples are useful because parameter
estimates from these samples tend to vary less than parameter estimates
from samples that are not spatially balanced (Barabesi and Franceschi,
2011; Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson
et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The first
spatially balanced sampling algorithm that saw widespread use was the
Generalized Random Tessellation Stratified (GRTS) algorithm (Stevens and
Olsen, 2004). To quantify the spatial balance of a sample, Stevens and
Olsen (2004) proposed loss metrics based on Voroni polygons. Since GRTS
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use Generalized Random Tessellation Stratified
(GRTS) sampling because it has several attractive properties: GRTS
sampling accommodates finite and infinite sampling frames; accommodates
equal, unequal, and proportional (to) size inclusion probabilities;
accommodates legacy (historical) sampling; accommodates a minimum
distance between units in a sample; accommodates reverse hierarchically
ordered replacement units in a sample (replacement units are units
available to be sampled if an original unit cannot be sampled); and is
available in the spsurvey R package Dumelle et al. (2021).

The GRTS algorithm samples from finite and infinite populations by
utilizing a mapping between two-dimensional and one-dimensional space.
The units in the two-dimensional sampling frame are divided into cells
using a hierarchical address. This hierarchical address is then used to
map the units from two-dimensional space to a one-dimensional line where
each unit's line length equals its inclusion probability. A systematic
sample is conducted on the line and linked back to a unit in
two-dimensional space, which results in the desired sample. Stevens and
Olsen (2004) and Dumelle et al. (2021) provide further details.

After selecting a GRTS sample, data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean is obtained by dividing
\(\hat{\tau}_{ht}\) by the population size.

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
these estimators rely on calculating \(\pi_{ij}\), the probability that
unit \(i\) and unit \(j\) are both in the sample -- this quantity can be
challenging if not impossible to calculate analytically. Second, these
estimators ignore the spatial locations of the units in the sampling
frame. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neigborhood
variance estimator tends reduce \(\text{Var}(\hat{\tau})\) compared to
variance estimators ignoring spatial locations, yielding narrower
confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of basing inference off of a specific
sampling design, we assume the data are generated by a spatial process.
Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be a response vector at locations \(s_1\), \(s_2\), . . . , \(s_N\) that
can be measured at the \(N\) population units and is represented as an
\(N \times 1\) vector. Suppose we want to predict some linear function
of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

We often only have a sample of the \(N\) population units. Denoting
quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with a subscript \emph{u},

@@

In addition to assuming the expectation of \(\bm{\delta}\) equals
\(\mathbf{0}\), we also assume that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{ps}\) is the partial sill measuring
coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the nugget
measuring fine-scale (indepenent) variability, \(\phi\) is the range
parameter measuring the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). Any spatial covariance function could be used in the place of the
exponential, however, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in (Ver Hoef, 2008), we
note here that the predictor and its variance are both moment-based,
meaning they don't rely on any distributional assumptions.

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses of a finite population. We
choose to use FPBK because it is faster than a Bayesian approach and
random forest and because Ver Hoef and Temesgen (2013) showed that the
method outperforms k-nearest-neighbors in many scenarios.

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & IRS & GRTS \\ 
  \hline
IRS & IRS-Design & GRTS-Design \\ 
  GRTS & IRS-Model & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.} 
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|lll}
   \hline
Sample Size (n) & 50 & 100 & 200 \\ 
  Layout & Random & Gridded & - \\ 
  Proportion of Dependent Error & 0 & 0.5 & 0.9 \\ 
  Response Type & Normal & Lognormal & - \\ 
   \hline
\end{tabular}
\caption{\label{tab:parmtab} Simulation parameters. Total variability for all scenarios was 2 so that the partial sill was 0, 1, or 1.8.} 
\end{table}

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T.-l., 2010. Geostatistical inference
under preferential sampling. Journal of the Royal Statistical Society:
Series C (Applied Statistics) 59, 191--232.

### 2021-10-12 inst/manuscript/notes_Madsen.Rmd
section 2 updates

>>>>

__Lisa M.__: Need a more precise definition of "miniature" in this context, and need an example.

__Lisa M.__: List major attractive properties, and detailed by @stevens2004spatially and @dumelle2021spsurvey.

__Lisa M.__ : Include formulas. Perhaps, but, these are very heavy in notation and matrix algebra. We might consider, however, adding the formulas to an Appendix.

### 2021-10-12 inst/manuscript/notes_VerHoef.Rmd
section 2 updates

>>>>

__Jay VH__: What about the design for model-based inference? Strictly speaking, it is fixed – there is no probabilistic use of a randomized design. However, we are going to have to deal with Diggle et al. (2010).

__Jay VH__: Saying "the distribution of the sampled population units mirrors the density of..." is confusing to me. Are these formal statistical definitions of distribution (cumulative distribution function) and density (probability density function)? Wouldn’t IRS sample be a miniature, as it should, on average, mirror a population?
